---
title: "MAST30027 Modern Applied, Statistics Assignment2"
date: "September 15,2019"
output:
  pdf_document: default
  word_document: default

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Name: Tianyi Mo

Student ID: 875556

Tutorial time: Tue 2.15pm

Tutor: Qiuyi Li


#1.Introduction
##1.1 Background
This report is about evaluating chimpanzees prosocial tendency by analysing and fitting model to the data.
The experiment has has two options, one is prosocial option, when human students participate the experiment, they nearly always choose the prosocial option when another student sits on the opposite side of the table. The question is whether a focal chimpanzee behaves similarly, choosing the prosocial option more often when another animal is present.

This report will analysis the dataset, visualize it, fit models to it and finally gives conclusion.


##1.2 Data
There are four attributtes in the raw data. By looking through the dataset, it is found that there are 7 chimpanzees and every chimpanzee has the same number of instance if data (72 for each chimpanzee), therefore it is not tend to bias towards particular a chimpanzee. It should also be noted the the data is also balanced in condition and prosoc_left attribute, which means that each chimpanzee has 16 experiment on each combination of condition and prosoc left.

These are the 4 attributes in the raw dataset.
actor (1 to 7)
condition (0 or 1):
prosoc left (0 or 1)
pulled left (0 or 1)

```{r}
# Load the dataset
dataset = read.delim("assign2.txt", header = TRUE, sep = " ")
```



#2. Preprocessing
##2.1 Create an New Attribute
Since the question is whether a focal chimpanzee choosing the prosocial option more often when another animal is present. Therefore, an new attribute "prosocial_action" is created, it means whether prosoc_left and pulled_left are same. If these two atributes are same, chimpanzee perform prosocially and prosocial_action has value TRUE and otherwise it has value FALSE.
```{r}
dataset['prosocial_action'] = (dataset$prosoc_left == dataset$pulled_left)
#print out first10 rows in the dataset after adding new attribute prosocial_action
dataset[1:10,]
```




#3.Visualization

##3.1Percentage of prosocial action and other factors
These plots shows how does the percentage of pro_social action depend on condition, actor and prosoc_left.
```{r}
# select the prosocial_action = TRUE data only.
x = xtabs( ~ condition + prosoc_left+ actor, data = dataset[dataset$prosocial_action ==TRUE,])
x=data.frame(x)

# divid the Freq(count) by total(=18) and plot
x["Pecentage_of_prosocial_action"] = x$Freq/18
```


#3.1.1 Percentage of prosocial action and Condition
```{r fig5, fig.height = 3, fig.width = 2}
#Mean of prosocial action and condition
prosocial = xtabs(prosocial_action ~ condition, data = dataset)
(prosocial_percentage = transform(prosocial, Freq=Freq/252))

#Boxplot of prosocial action and condition
plot(Pecentage_of_prosocial_action~condition,x)
```

The first plot shows the relationship between condition and prosocial_action. The median decrease, and the variance increase. The mean percentage of choose prosocial increase from 0.5555556 to 0.5793651 in condition 1.
On average, chimapnzees choose the prosocial option sightly more frequently in condition 1 (when there is
another chimpanzee in opposite). Since the mean incrases and the median decrease, the relationship between them is not clear.

#3.1.2 Percentage of prosocial action and Actor
```{r}
# It can be seen form the data that the mean prosocial_action for each actor are diffenent. 
prosocial = xtabs(prosocial_action ~ actor, data = dataset)
(prosocial_percentage = transform(prosocial, Freq=Freq/72))

#Boxplot of prosocial action and condition
plot(Pecentage_of_prosocial_action~actor,x)
```
The second plot shows that each actor has different response in the test. The prosocial_action is depend on actor. For example, the second actor has the largest variance and actor 1 and 5 have relatively high median and low variance. The trend is meaningless in this plot because actor is not ordinal class.



#3.1.3 Percentage of prosocial action and Prosoc_left
```{r}
prosocial = xtabs(prosocial_action ~ prosoc_left, data = dataset)
(prosocial_percentage = transform(prosocial, Freq=Freq/252))

#Boxplot of prosocial action and prosoc_left
plot(Pecentage_of_prosocial_action~prosoc_left,x)
```
The third plot shows the relationship between prosoc_left and prosocial_action. It can be seen the left and right has different result on prosocial_action. When prosocial option is at right, chimpanzees has 0.4880952 to choose it, but when it is at left, chimpanzees has 0.6468254 probability to choose it.

\newpage
#3.2 Interactions Between Two Factors
```{r fig2, fig.height = 3.7, fig.width = 6}
with(dataset, interaction.plot(condition,actor,prosocial_action))
with(dataset, interaction.plot(actor,condition,prosocial_action))
```
```{r fig3, fig.height = 3.7, fig.width = 6}
with(dataset, interaction.plot(actor,prosocial_action,condition))
```
#It seems that there are some interaction between actor and prosocial_action and all actors have different mean of pro_action.


##3.3 Outlier Detection
From the previous plots, the second chimpanzee (actor 2) show strange result. In both conditions, it chooses proaction option at exactly at 0.5. By looking back to the raw dataset. It can be found that the it always choose the left one regardless of any other factors. There is no similar behaviour like this among other chimpanzees and it might be an outlier.

```{r}
# Second chimpanzee always pulled the left lever
xtabs(~prosoc_left+pulled_left, data =  dataset[dataset$actor == 2,])
```

```{r}
# Other chimpanzees pulled both left and right lever
# for example chimpanzees 1
xtabs(~prosoc_left+pulled_left, data =  dataset[dataset$actor == 1,])
```

This might be an error in data collection process and might be nomal behaviour of chimpanzee species. Based on the data, I cannot decided which one it should be. This problem should be checked with the zoologist and experts with domain knowledges to decided whether to remove data from chimpanzee 2.


\newpage

\hfill\break

##4.Contengency Table
Use contengency table to test the dependency of prosocial_action, condition.
Using three way contengency table to test the mutually independency of the 3 factors. Poisson regression is used because poisson regression with log link will gives chi-sqare distribution.

H0 = prosocial_action & condition & actor are independent

H1 = prosocial_action & condition & actor are not independent
```{r}
x = xtabs( ~ condition + prosocial_action, data = dataset)
(x = data.frame(x))
```
The cell counts need to be at least 5 for the deviance to has chi-square distribution as describe in the lectue. From the data count above, it can be seen that the minimum count is 13.

```{r}
#fit poisson model
model0 = glm(Freq ~ condition + prosocial_action, family = poisson, data = x)
#deviance of the poission regression model
deviance(model0)
#degrees of freedom
df.residual(model0)
# calculate the p-value
pchisq(deviance(model0), df = df.residual(model0), lower.tail = FALSE)

```
The p-value is 0.9598556, it is much larger than 0.05, thus we cannot reject H0 and prosocial_action & condition & action are independent. Since we are interested in whether prosocial_action is related the condition (whether there is another chimpanzee in another side), from the result we can say that prosocial_action is independent of condition. And chimpanzee don't behave similarly as human.

\newpage
#5. Binomial Regression
#5.1 Fitting data with binomial regression
If we let the count be response variable, and all four factors (condition, prosoc_left, pulled_left, actor) as explanation variable. The count has binomial distribution with n equals to 18 and p is the probability of choosing the prosocial_action.
Since the n is only 18 and p is large, we cannot use poisson distribution to approximate the binomial distribution.

##5.1.1 Fit binomial model
```{r}

x = xtabs( ~ condition + prosoc_left + pulled_left+ actor, data = dataset)
x = data.frame(x)
#set last column to n = 18
x["total"] = 18
(x = x[x$prosoc_left == x$pulled_left,])

#the data have 28 rows
dim(x)

#fit binomial regression model
fullmodel = glm(cbind(total-Freq,Freq) ~ (condition + prosoc_left +actor)^2, family = binomial, data = x)
```

##5.1.2 Model selsction
```{r}
#use AIC to select model
finalmodel = step(fullmodel)
```
The fianl model after model selection is y ~ prosoc_left + actor + prosoc_left:actor
```{r}
summary(finalmodel)
deviance(finalmodel)
df.residual(finalmodel)
pchisq(deviance(finalmodel), df = df.residual(finalmodel), lower.tail = FALSE)
anova(finalmodel,fullmodel,test = "Chisq")
```

The model comparision test between fullmodel and final model gives p-value = 0.8402, thus the final model is adequate.

The fianl model after model selection is y ~ prosoc_left + actor + prosoc_left:actor. Therefore the number of times choosing prosocial action does dependent on condition because it is not in the final model.

##5.1.3 Check overdispersion
```{r}
(phihat = sum(residuals(finalmodel,type = "pearson")^2)/finalmodel$df.residual)
```
The phihat is 0.9971355. It is close to 1 and there is no overdispersion or underdispersion exist.

##5.1.4 Diagonistic plot
```{r}
library(faraway) # for halfnorm function
D_res <- residuals(finalmodel)
P_res <- residuals(finalmodel, type="pearson")
lever <- influence(finalmodel)$hat
J_res <- rstudent(finalmodel)
Cooks <- cooks.distance(finalmodel)
eta_hat <- predict(finalmodel, type="link")
par(mfrow=c(2,2))
par(mar=c(4,4,1,2))
plot(eta_hat, D_res, ylab="dev res")
lines(predict(loess(D_res ~ eta_hat)), col="red")
halfnorm(D_res, ylab="dev res")
halfnorm(J_res, ylab="jackknife res")
halfnorm(Cooks, ylab="Cooks dist")
```

The first plot has some data points have extremely small and large eta_hat values. From the eta_hat data, it can be found that it is due to the second actor always choose the left lever. In section 3.3, it is mentioned that actor 2 might be an outlier, and we can remove it and see whether it influence our model.

Beside this, all other points have reasonable cooks distance and residual.

\newpage

##5.2 After remove the outlier, fitting data with binomial regression

```{r}
#remove actor2
x = xtabs( ~ condition + prosoc_left + pulled_left+ actor, data = dataset[dataset$actor !=2,])
x = data.frame(x)
#set last column to n = 18

x["total"] = 18
x = x[x$prosoc_left == x$pulled_left,]

#the data have 24 rows
dim(x)

#fit binomial regression model
fullmodel = glm(cbind(total-Freq,Freq) ~ (condition + prosoc_left +actor)^2, family = binomial, data = x)

#use AIC to select model
finalmodel = step(fullmodel)
```
The model after remove outlier does not change. It is still y ~ prosoc_left + actor + prosoc_left:actor.
```{r}
(phihat = sum(residuals(finalmodel,type = "pearson")^2)/finalmodel$df.residual)

library(faraway) # for halfnorm function
D_res <- residuals(finalmodel)
P_res <- residuals(finalmodel, type="pearson")
lever <- influence(finalmodel)$hat
J_res <- rstudent(finalmodel)
Cooks <- cooks.distance(finalmodel)
eta_hat <- predict(finalmodel, type="link")
par(mfrow=c(2,2))
par(mar=c(4,4,1,2))
plot(eta_hat, D_res, ylab="dev res")
lines(predict(loess(D_res ~ eta_hat)), col="red")
halfnorm(D_res, ylab="dev res")
halfnorm(J_res, ylab="jackknife res")
halfnorm(Cooks, ylab="Cooks dist")


deviance(finalmodel)
df.residual(finalmodel)
pchisq(deviance(finalmodel), df = df.residual(finalmodel), lower.tail = FALSE)
anova(finalmodel,fullmodel,test = "Chisq")
```
The removal of outlier dose not change out model, the model is still y ~ prosoc_left + actor + prosoc_left:actor, although our p-value decrease from 0.8402 to 0.7584. Since the model does not change, we can sill conclude that the number of times choosing prosocial action does dependent on condition.

##6.Conclusion
In this report, section 2 add an new attribute prosocial_action, section 3 plot the relationsip and interactions between factors, section 4 show the prosocial action and condition are independent using contengencytable and section 5 show the number of times choosing prosocial action only dependent on actor and prosoc_left. Thus focal chimpanzee dose not behave like human that choose the prosocial option more often when another animal is present.

As described in section 3.3, actor 2 might be an outlier because it always choose the left lever and is not influence by any other factor. This require some domain knowledge to decided whether it should be removed. However, whether it is removed or not dose not change the result.
